---
title: "latentcor"
author: "Mingze Huang, Irina Gaynanova, Christian L. Müller"
date: "`r Sys.Date()`"
bibliography: latentcor.bib
output: rmarkdown::html_vignette
extra_dependencies: ["amsmath"]
nocite: |
  @croux2013robust
  @filzmoser2021pcapp
  @liu2009nonparanormal
  @fox2019poly
vignette: >
  %\VignetteIndexEntry{latentcor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(latentcor)
```


# Introduction

R package `latentcor` utilizes the powerful semi-parametric latent Gaussian copula models to estimate latent correlations between mixed data types. The package allows to estimate correlations between any of continuous/binary/ternary/zero-inflated (truncated) variable types. The underlying implementation takes advantage of fast multi-linear interpolation scheme with a clever choice of grid points that give the package a small memory footprint, and allows to use the latent correlations with sub-sampling and bootstrapping.

# Getting started

## A simple example with two variables

First, we will generate a pair of variables with different types using a sample size $n=100$ which will serve as example data. Here first variable will be ternary, and second variable will be continuous.

```{r data_generation}
simdata = GenData(n = 100, types = c("ter", "con"))
```

The output of `GenData` is a list with 2 elements:

```{r data_output}
names(simdata)
```

  - `X`: a matrix ($100\times 2$), the first column is the ternary variable; the second column is the continuous variable.
  
```{r data_matrix}
X = simdata$X
head(X, n = 6L)
```

  - `plotX`: NULL (`showplot = FALSE`, can be changed to display the plot of generated data in`GenData` input).

```{r data_plot}
simdata$plotX
```

Then we can estimate the latent correlation matrix based on these 2 variables using `estR` function.

```{r estimation}
estimate = estR(X, types = c("ter", "con"))
```

The output of `estR` is a list with several elements:

```{r estimation_output}
names(estimate)
```

  - `zratios` is a list has the same length as the number of variables. Here the first element is a ($2\times1$) vector indicating the cumulative proportions for zeros and ones in the ternary variable (e.g. first element in vector is the proportion of zeros, second element in vector is the proportion of zeros and ones.) The second element of the list is NA for continuous variable.

```{r zratios}
estimate$zratios
```

  - K: Kendall $\tau$ ($\tau_{a}$) correlation matrix for these 2 variables. 
  
```{r Kendall}
estimate$K
```  

R: estimated latent correlation matrix for these 2 variable.

```{r latent_correlation}
estimate$R
``` 

plotR: NULL by default as `showplot = FALSE` in `estR`. Otherwise displays a heatmap of latent correlation matrix.

```{r heatmap}
estimate$plotR
``` 


## Another example

# Estimation methods

## Original method

### Latent Gaussian Copula Model for Mixed Data

`latentcor` utilizes the powerful semi-parametric latent Gaussian copula models to estimate latent correlations between mixed data types (continuous/binary/ternary/truncated or zero-inflated). Below we review the definitions for each type.

***Definition of continuous model***

A random $X\in\cal{R}^{p}$ satisfies the Gaussian copula (or nonparanormal) model if there exist monotonically increasing $f=(f_{j})_{j=1}^{p}$ with $Z_{j}=f_{j}(X_{j})$ satisfying $Z\sim N_{p}(0, \Sigma)$, $\sigma_{jj}=1$; we denote $X\sim NPN(0, \Sigma, f)$.

```{r continuous}
X = GenData(types = "con")$X
head(X, n = 6L)
```

***Definition of binary model***

A random $X\in\cal{R}^{p}$ satisfies the binary latent Gaussian copula model if there exists $W\sim NPN(0, \Sigma, f)$ such that $X_{j}=I(W_{j}>c_{j})$, where $I(\cdot)$ is the indicator function and $c_{j}$ are constants.

```{r binary}
X = GenData(types = "bin")$X
head(X, n = 6L)
```

***Definition of ternary model***

A random $X\in\cal{R}^{p}$ satisfies the ternary latent Gaussian copula model if there exists $W\sim NPN(0, \Sigma, f)$ such that $X_{j}=I(W_{j}>c_{j})+I(W_{j}>c'_{j})$, where $I(\cdot)$ is the indicator function and $c_{j}<c'_{j}$ are constants.

```{r ternary}
X = GenData(types = "ter")$X
head(X, n = 6L)
```

***Definition of truncated or zero-inflated model***

A random $X\in\cal{R}^{p}$ satisfies the truncated latent Gaussian copula model if there exists $W\sim NPN(0, \Sigma, f)$ such that $X_{j}=I(W_{j}>c_{j})W_{j}$, where $I(\cdot)$ is the indicator function and $c_{j}$ are constants.

```{r truncated}
X = GenData(types = "tru")$X
head(X, n = 6L)
```


***Mixed latent Gaussian copula model***

The mixed latent Gaussian copula model jointly models $W=(W_{1}, W_{2}, W_{3}, W_{4})\sim NPN(0, \Sigma, f)$ such that $X_{1j}=W_{1j}$, $X_{2j}=I(W_{2j}>c_{2j})$, $X_{3j}=I(W_{3j}>c_{3j})+I(W_{3j}>c'_{3j})$  and $X_{4j}=I(W_{4j}>c_{4j})W_{4j}$.

```{r mixed}
X = GenData(types = c("con", "bin", "ter", "tru"))$X
head(X, n = 6L)
```

***Kendall's $\tau$ ($\tau_{a}$)***

The estimation of latent correlation matrix $\Sigma$ is achieved via the bridge function $F$ which is defined such that $E(\hat{\tau}_{jk})=F(\sigma_{jk})$, where $\sigma_{jk}$ is the latent correlation between variables $j$ and $k$, and $\hat{\tau}_{jk}$ is the corresponding sample Kendall's $\tau$. Given observed $\mathbf{x}_{j}, \mathbf{x}_{k}\in\cal{R}^{n}$,

$$
\hat{\tau}_{jk}=\hat{\tau}(\mathbf{x}_{j}, \mathbf{x}_{k})=\frac{2}{n(n-1)}\sum_{1\le i<i'\le n}sign(x_{ij}-x_{i'j})sign(x_{ik}-x_{i'k}),
$$
where $n$ is the sample size.

```{r KendallTau}
estR(X, types = c("con", "bin", "ter", "tru"), method = "original")$K
```

Using $F$, a moment-based estimator is $\hat{\sigma}_{jk}=F^{-1}(\hat{\tau}_{jk})$ with the corresponding  $\hat{\Sigma}$ being consistent for $\Sigma$ [@fan2017high; @quan2018rank; @yoon2021fast]. The explicit form of $F$ has been derived for all combinations of continuous(C)/binary(B)/ternary(N)/truncated(T) variable types [@fan2017high; @yoon2021fast]. 

### Theorem of bridge functions

Let $W_{1}\in\cal{R}^{p_{1}}$, $W_{2}\in\cal{R}^{p_{2}}$, $W_{3}\in\cal{R}^{p_{3}}$, $W_{4}\in\cal{R}^{p_{4}}$ be such that $W=(W_{1}, W_{2}, W_{3}, W_{4})\sim NPN(0, \Sigma, f)$ with $p=p_{1}+p_{2}+p_{3}+p_{4}$. Let $X=(X_{1}, X_{2}, X_{3}, X_{4})\in\cal{R}^{p}$ satisfy $X_{j}=W_{j}$ for $j=1,...,p_{1}$, $X_{j}=I(W_{j}>c_{j})$ for $j=p_{1}+1, ..., p_{1}+p_{2}$, $X_{j}=I(W_{j}>c_{j})W_{j}$ for $j=p_{1}+p_{2}+1, ..., p_{3}$ and $X_{j}=I(W_{j}>c_{j})+I(W_{j}>c'_{j})$ for $j=p_{1}+p_{2}+p_{3}+1, ..., p$ with $\Delta_{j}=f(c_{j})$. The rank-based estimator of $\Sigma$ based on the observed $n$ realizations of $X$ is the matrix $\mathbf{\hat{R}}$ with $\hat{r}_{jj}=1$, $\hat{r}_{jk}=\hat{r}_{kj}=F^{-1}(\hat{\tau}_{jk})$ with block structure

$$
\mathbf{\hat{R}}=\begin{pmatrix}
F_{CC}^{-1}(\hat{\tau}) & F_{CB}^{-1}(\hat{\tau}) & F_{CT}^{-1}(\hat{\tau})\\
F_{BC}^{-1}(\hat{\tau}) & F_{BB}^{-1}(\hat{\tau}) & F_{BT}^{-1}(\hat{\tau})\\
F_{TC}^{-1}(\hat{\tau}) & F_{TB}^{-1}(\hat{\tau}) & F_{TT}^{-1}(\hat{\tau})
\end{pmatrix}
$$

|Type | continuous | binary | zero-inflated (truncated) | ternary |
|-----|----------|----------|----------|----------|
|continuous | @liu2009nonparanormal |- | -| - |
|binary | @fan2017high | @fan2017high | - | - |
|zero-inflated (truncated) | @yoon2020sparse | @yoon2020sparse | @yoon2020sparse | - |
|ternary | @quan2018rank | @quan2018rank | See Appendix | @quan2018rank |

$$
F(\cdot)=\begin{cases}
\frac{2}{\pi}sin^{-1}(r)  &  for \; CC\\
4\Phi_{2}(\Delta_{j},0;\frac{r}{\sqrt{2}})-2\Phi(\Delta_{j}))  &  for \; BC\\
2[\Phi_{2}(\Delta_{j},\Delta_{k};r)-\Phi(\Delta_{j})\Phi(\Delta_{k})]  &  for \; BB\\
-2\Phi_{2}(-\Delta_{j},0;\frac{1}{\sqrt{2}})+4\Phi_{3}(-\Delta_{j},0,0;\Sigma_{3}(r))  &  for \; TC\\
2[1-\Phi(\Delta_{j})]\Phi(\Delta_{k})-2\Phi_{3}(-\Delta_{j},\Delta_{k},0;\Sigma_{3a}(r))\\
\;\;\;-2\Phi_{3}(-\Delta_{j},\Delta_{k},0;\Sigma_{3b}(r))  &  for \; TB\\
-2\Phi_{4}(-\Delta_{j},-\Delta_{k},0,0;\Sigma_{4a}(r))\\
\;\;\;+2\Phi_{4}(-\Delta_{j},-\Delta_{k},0,0;\Sigma_{4b}(r))  &  for \; TT\\
4\Phi_{2}(\Delta_{j}^{2},0;\frac{r}{\sqrt{2}})-2\Phi(\Delta_{j}^{2})\\
\;\;\;+4\Phi_{3}(\Delta_{j}^{1},\Delta_{j}^{2},0;\Sigma_{3c}(r))-2\Phi(\Delta_{j}^{1})\Phi(\Delta_{j}^{2})  &  for \; NC\\
2\Phi_{2}(\Delta_{j}^{2},\Delta_{k},r)(1-\Phi(\Delta_{j}^{1}))\\
\;\;\;-2\Phi(\Delta_{j}^{2})(\Phi(\Delta_{k})-\Phi_{2}(\Delta_{j}^{1},\Delta_{k},r))  &  for \; NB\\
-2\Phi(-\Delta_{j}^{1})\Phi(\Delta_{j}^{2}) + 2\Phi_{3}(-\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k};\Sigma_{3e}(r))\\
\;\;\;+2\Phi_{4}(-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\Sigma_{4c}(r))\\
\;\;\;+2\Phi_{4}(-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\Sigma_{4d}(r))  &  for \; NT\\
2\Phi_{2}(\Delta_{j}^{2},\Delta_{k}^{2};r)\Phi_{2}(-\Delta_{j}^{1},-\Delta_{k}^{1};r)\\
\;\;\;-2[\Phi(\Delta_{j}^{2})-\Phi_{2}(\Delta_{j}^{2},\Delta_{k}^{1};r)][\Phi(\Delta_{k}^{2}-\Phi_{2}(\Delta_{j}^{1},\Delta_{k}^{2};r))]  &  for \; NN
\end{cases}
$$

where $\Delta_{j}=\Phi^{-1}(\pi_{0j})$, $\Delta_{k}=\Phi^{-1}(\pi_{0k})$, $\Delta_{j}^{1}=\Phi^{-1}(\pi_{0j})$, $\Delta_{j}^{2}=\Phi^{-1}(\pi_{0j}+\pi_{1j})$, $\Delta_{k}^{1}=\Phi^{-1}(\pi_{0k})$, $\Delta_{k}^{2}=\Phi^{-1}(\pi_{0k}+\pi_{1k})$,

$$
\Sigma_{3}(r)=
\begin{pmatrix}
1 & \frac{1}{\sqrt{2}} & \frac{r}{\sqrt{2}}\\
\frac{1}{\sqrt{2}} & 1 & r \\
\frac{r}{\sqrt{2}} & 2 & 1
\end{pmatrix}, \;\;\;
\Sigma_{3a}(r)=
\begin{pmatrix}
1 & -r & \frac{1}{\sqrt{2}} \\
-r & 1 & -\frac{r}{\sqrt{2}} \\
\frac{1}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & 1
\end{pmatrix},
$$

$$
\Sigma_{3b}(r)=
\begin{pmatrix}
1 & 0 & -\frac{1}{\sqrt{2}} \\
0 & 1 & -\frac{r}{\sqrt{2}} \\
-\frac{1}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & 1
\end{pmatrix}, \;\;\;
\Sigma_{3c}(r)=
\begin{pmatrix}
1 & 0 & \frac{r}{\sqrt{2}} \\
0 & 1 & -\frac{r}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & 1
\end{pmatrix},
$$
$$
\Sigma_{3d}(r)=
\begin{pmatrix}
1 & 0 & -r \\
0 & 1 & 0 \\
-r & 0 & 1
\end{pmatrix}, \;\;\;
\Sigma_{3e}(r)=
\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & r \\
0 & r & 1
\end{pmatrix},
$$
$$
\Sigma_{4a}(r)=
\begin{pmatrix}
1 & 0 & \frac{1}{\sqrt{2}} & -\frac{r}{\sqrt{2}} \\
0 & 1 & -\frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} \\
\frac{1}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & 1 & -r \\
-\frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} & -r & 1
\end{pmatrix}, \;\;\;
\Sigma_{4b}(r)=
\begin{pmatrix}
1 & r & \frac{1}{\sqrt{2}} & \frac{r}{\sqrt{2}} \\
r & 1 & \frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} \\
\frac{1}{\sqrt{2}} & \frac{r}{\sqrt{2}} & 1 & r \\
\frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} & r & 1
\end{pmatrix},
$$
$$
\Sigma_{4c}(r)=
\begin{pmatrix}
1 & 0 & 0 & \frac{r}{\sqrt{2}} \\
0 & 1 & -r & \frac{r}{\sqrt{2}} \\
0 & -r & 1 & -\frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 1
\end{pmatrix} \;\;\;\text{and}\;\;\;
\Sigma_{4d}(r)=
\begin{pmatrix}
1 & 0 & r & \frac{r}{\sqrt{2}} \\
0 & 1 & 0 & \frac{r}{\sqrt{2}} \\
r & 0 & 1 & \frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} & 1
\end{pmatrix}.
$$

***Algorithm for Original method `latentcor::estR(X, method = "original")`***

**Input**: $F(r)=F(r, \mathbf{\Delta})$ - bridge function based on the type of variables $j$, $k$

   - Step 1. Calculate $\hat{\tau}_{jk}$ using (1).
   
   - Step 2. For binary/truncated variable $j$, set $\hat{\mathbf{\Delta}_{j}}=\hat{\Delta}_{j}=\Phi^{-1}(\pi_{0j})$ with $\pi_{0j}=\sum_{i=1}^{n}\frac{I(x_{ij}=0)}{n}$. For ternary variable $j$, set $\hat{\mathbf{\Delta}_{j}}=(\hat{\Delta}_{j}^{1}, \hat{\Delta_{j}^{2}})$ where $\hat{\Delta}_{j}^{1}=\Phi^{-1}(\pi_{0j})$ and $\hat{\Delta}_{j}^{2}=\Phi^{-1}(\pi_{0}+\pi_{1})$ with $\pi_{0j}=\sum_{i=1}^{n}\frac{I(x_{ij}=0)}{n}$ and $\pi_{1j}=\sum_{i=1}^{n}\frac{I(x_{ij}=1)}{n}$.
   
   - Compute $F^{-1}(\hat{\tau}_{jk})$ as $\hat{r}_{jk}=argmin\{F(r)-\hat{\tau}_{jk}\}^{2}$ solved via `optimize` function in *R*.


## Approximation method

### Inversion via Multilinear Interpolation

The inverse bridge function is an analytic function of at most 5 parameters (see Theorem 1):

  - Kendall's $\tau$
  - Proportion of zeros in the 1st variable 
  - (Possibly) proportion of zeros and ones in the 1st variable
  - (Possibly) proportion of zeros in the 2nd variable
  - (Possibly) proportion of zeros and ones in the 2nd variable


In short, d-dimensional multilinear interpolation uses a weighted average of $2^{d}$ neighbors to approximate the function values at the points within the d-dimensional cube of the neighbors [@yoon2021fast]. This can be done by R package `chebpol` [@R-chebpol].

### Approximation via hybrid Scheme 

***Algorithm for approximation method `latentcor::estR(X, method = "approx")`***

**Input**: Let $\tilde{g}=h(g)$, pre-computed values $F^{-1}(h^{-1}(\tilde{g}))$ on a fixed grid $\tilde{g}\in\tilde{\cal{G}}$ based on the type of variables $j$ and $k$. For binary/continuous case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j})$; for binary/binary case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}, \tilde{\Delta}_{k})$; for truncated/continuous case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j})$; for truncated/truncated case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}, \tilde{\Delta}_{k})$; for ternary/continuous case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}^{1}, \tilde{\Delta}_{j}^{2})$; for ternary/binary case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}^{1}, \tilde{\Delta}_{j}^{2}, \tilde{\Delta}_{k})$; for ternary/truncated case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}^{1}, \tilde{\Delta}_{j}^{2}, \tilde{\Delta}_{k})$; for ternay/ternary case, $\tilde{g}=(\tilde{\tau}_{jk}, \tilde{\Delta}_{j}^{1}, \tilde{\Delta}_{j}^{2}, \tilde{\Delta}_{k}^{1}, \tilde{\Delta}_{k}^{2})$.

  - Step 1 and Step 2 same as Algorithm 1.
  
  - Step 3. If $|\hat{\tau}_{jk}|\le ratio\times \bar{\tau}(\cdot)$, apply Algorithm 3; Otherwise apply Algorithm 1.

To avoid interpolation in areas with high approximation errors close to the boundary, we use hybrid scheme [@yoon2021fast]. The derivation of approximate bound for BC, BB, TC, TB, TT cases see [@yoon2021fast]. The derivation of approximate bound for NC, NB, NN, NT case see Appendix.

$$
\bar{\tau}(\cdot)=
\begin{cases}
2\pi_{0j}(1-\pi_{0j})  &   for \; BC \; case\\
2\min(\pi_{0j},\pi_{0k})\{1-\max(\pi_{0j}, \pi_{0k})\}  &   for \; BB \; case\\
1-(\pi_{0j})^{2}  &   for \; TC \; case\\
2\max(\pi_{0k},1-\pi_{0k})\{1-\max(\pi_{0k},1-\pi_{0k},\pi_{0j})\}  &   for \; TB \; case\\
1-\{\max(\pi_{0j},\pi_{0k})\}^{2}  &   for \; TT \; case\\
2\{\pi_{0j}(1-\pi_{0j})+\pi_{1j}(1-\pi_{0j}-\pi_{1j})\}  &   for \; NC \; case\\
2\min(\pi_{0j}(1-\pi_{0j})+\pi_{1j}(1-\pi_{0j}-\pi_{1j}),\pi_{0k}(1-\pi_{0k}))  &   for \; NB \; case\\
1-\{\max(\pi_{0j},\pi_{1j},1-\pi_{0j}-\pi_{1j},\pi_{0k})\}^{2}  &   for \; NT \; case\\
2\min(\pi_{0j}(1-\pi_{0j})+\pi_{1j}(1-\pi_{0j}-\pi_{1j}), \\
\;\;\;\;\;\;\;\;\;\;\pi_{0k}(1-\pi_{0k})+\pi_{1k}(1-\pi_{0k}-\pi_{1k}))  &   for \; NN \; case
\end{cases}
$$


**Rescale Grid for Interpolation**

Note that $|\hat{\tau}|\le \bar{\tau}$, the grid does not need to cover the domain $\tau\in[-1, 1]$. Instead, we rescale them as following:
$\tilde{\tau}_{jk}=\frac{\tau_{jk}}{\bar{\tau}_{jk}}\in[-1, 1]$, where $\bar{\tau}_{jk}$ applies the approximation bound function with respect to the data types corresponding to variable $j$ and $k$. For ternary variable $j$, we know $\Delta_{j}^{2}>\Delta_{j}^{1}$ always holds since $\Delta_{j}^{1}=\Phi^{-1}(\pi_{0j})$ and $\Delta_{j}^{2}=\Phi^{-1}(\pi_{0j}+\pi_{1j})$, the grid should not cover the domain for the areas of $\Delta_{j}^{2}\ge\Delta_{j}^{1}$. So that we rescale them as following: $\tilde{\Delta}_{j}^{1}=\frac{\Delta_{j}^{1}}{\Delta_{j}^{2}}\in[0, 1]$; $\tilde{\Delta}_{j}^{2}=\Delta_{j}^{2}\in[0, 1]$.

# Appendix

## Derivation for bridge function for ternary/truncated case

Without loss of generality, let $j=1$ and $k=2$. By the definition of Kendall's $\tau$,
$$
    \tau_{12}=E(\hat{\tau}_{12})=E[\frac{2}{n(n-1)}\sum_{1\leq i\leq i' \leq n} sign\{(X_{i1}-X_{i'1})(X_{i2}-X_{i'2})\}]
$$
Since $X_{1}$ is ternary,
\begin{align}
    &sign(X_{1}-X_{1}') \nonumber\\ =&[I(U_{1}>C_{11},U_{1}'\leq C_{11})+I(U_{1}>C_{12},U_{1}'\leq C_{12})-I(U_{1}>C_{12},U_{1}'\leq C_{11})] \nonumber\\
    &-[I(U_{1}\leq C_{11}, U_{1}'>C_{11})+I(U_{1}\leq C_{12}, U_{1}'>C_{12})-I(U_{1}\leq C_{11}, U_{1}'>C_{12})] \nonumber\\
    =&[I(U_{1}>C_{11})-I(U_{1}>C_{11},U_{1}'>C_{11})+I(U_{1}>C_{12})-I(U_{1}>C_{12},U_{1}'>C_{12}) \nonumber\\
    &-I(U_{1}>C_{12})+I(U_{1}>C_{12},U_{1}'>C_{11})] \nonumber\\
    &-[I(U_{1}'>C_{11})-I(U_{1}>C_{11},U_{1}'>C_{11})+I(U_{1}'>C_{12})-I(U_{1}>C_{12},U_{1}'>C_{12}) \nonumber\\
    &-I(U_{1}'>C_{12})+I(U_{1}>C_{11},U_{1}'>C_{12})] \nonumber\\
    =&I(U_{1}>C_{11})+I(U_{1}>C_{12},U_{1}'>C_{11})-I(U_{1}'>C_{11})-I(U_{1}>C_{11},U_{1}'>C_{12}) \nonumber\\
    =&I(U_{1}>C_{11},U_{1}'\leq C_{12})-I(U_{1}'>C_{11},U_{1}\leq C_{12})
\end{align}
Since $X_{2}$ is truncated, $C_{1}>0$ and
\begin{align}
    sign(X_{2}-X_{2}')=&-I(X_{2}=0,X_{2}'>0)+I(X_{2}>0,X_{2}'=0) \nonumber\\
    &+I(X_{2}>0,X_{2}'>0)sign(X_{2}-X_{2}') \nonumber\\
    =&-I(X_{2}=0)+I(X_{2}'=0)+I(X_{2}>0,X_{2}'>0)sign(X_{2}-X_{2}')
\end{align}
Since $f$ is monotonically increasing, $sign(X_{2}-X_{2}')=sign(Z_{2}-Z_{2}')$,
\begin{align}
    \tau_{12}=&E[I(U_{1}>C_{11},U_{1}'\leq C_{12}) sign(X_{2}-X_{2}')] \nonumber\\ &-E[I(U_{1}'>C_{11},U_{1}\leq C_{12}) sign(X_{2}-X_{2}')] \nonumber\\
    =&-E[I(U_{1}>C_{11},U_{1}'\leq C_{12}) I(X_{2}=0)] \nonumber\\
    &+E[I(U_{1}>C_{11},U_{1}'\leq C_{12}) I(X_{2}'=0)] \nonumber\\
    &+E[I(U_{1}>C_{11},U_{1}'\leq C_{12})I(X_{2}>0,X_{2}'>0)sign(Z_{2}-Z_{2}')] \nonumber\\
    &+E[I(U_{1}'>C_{11},U_{1}\leq C_{12}) I(X_{2}=0)] \nonumber\\
    &-E[I(U_{1}'>C_{11},U_{1}\leq C_{12}) I(X_{2}'=0)] \nonumber\\
    &-E[I(U_{1}'>C_{11},U_{1}\leq C_{12})I(X_{2}>0,X_{2}'>0)sign(Z_{2}-Z_{2}')]  \nonumber\\
    =&-2E[I(U_{1}>C_{11},U_{1}'\leq C_{12}) I(X_{2}=0)] \nonumber\\
    &+2E[I(U_{1}>C_{11},U_{1}'\leq C_{12}) I(X_{2}'=0)] \nonumber\\
    &+E[I(U_{1}>C_{11},U_{1}'\leq C_{12})I(X_{2}>0,X_{2}'>0)sign(Z_{2}-Z_{2}')] \nonumber\\
    &-E[I(U_{1}'>C_{11},U_{1}\leq C_{12})I(X_{2}>0,X_{2}'>0)sign(Z_{2}-Z_{2}')]
\end{align}
From the definition of $U$, let $Z_{j}=f_{j}(U_{j})$ and $\Delta_{j}=f_{j}(C_{j})$ for $j=1,2$. Using $sign(x)=2I(x>0)-1$, we obtain
\begin{align}
    \tau_{12}=&-2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq \Delta_{12},Z_{2}\leq \Delta_{2})]+2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq \Delta_{12},Z_{2}'\leq \Delta_{2})] \nonumber\\
    &+2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq \Delta_{12})I(Z_{2}>\Delta_{2},Z_{2}'>\Delta_{2},Z_{2}-Z_{2}'>0)] \nonumber\\
    &-2E[I(Z_{1}'>\Delta_{11},Z_{1}\leq \Delta_{12})I(Z_{2}>\Delta_{2},Z_{2}'>\Delta_{2},Z_{2}-Z_{2}'>0)] \nonumber\\
    =&-2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq \Delta_{12}, Z_{2}\leq \Delta_{2})]+2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq \Delta_{12}, Z_{2}'\leq \Delta_{2})] \nonumber\\
    &+2E[I(Z_{1}>\Delta_{11},Z_{1}'\leq\Delta_{12},Z_{2}'>\Delta_{2},Z_{2}>Z_{2}')] \nonumber\\
    &-2E[I(Z_{1}'>\Delta_{11},Z_{1}\leq\Delta_{12},Z_{2}'>\Delta_{2},Z_{2}>Z_{2}')]
\end{align}
Since $\{\frac{Z_{2}'-Z_{2}}{\sqrt{2}}, -Z{1}\}$, $\{\frac{Z_{2}'-Z_{2}}{\sqrt{2}}, Z{1}'\}$ and $\{\frac{Z_{2}'-Z_{2}}{\sqrt{2}}, -Z{2}'\}$ are standard bivariate normally distributed variables with correlation $-\frac{1}{\sqrt{2}}$, $r/\sqrt{2}$ and $-\frac{r}{\sqrt{2}}$, respectively, by the definition of $\Phi_3(\cdot,\cdot, \cdot;\cdot)$ and $\Phi_4(\cdot,\cdot, \cdot,\cdot;\cdot)$ we have
\begin{align}
    F_{NT}(r;\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k})= & -2\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k};\begin{pmatrix}
1 & 0 & -r \\
0 & 1 & 0 \\
-r & 0 & 1
\end{pmatrix} \right\} \nonumber\\
    &+2\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k};\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & r \\
0 & r & 1
\end{pmatrix}\right\}\nonumber \\
    & +2\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & 0 & \frac{r}{\sqrt{2}} \\
0 & 1 & -r & \frac{r}{\sqrt{2}} \\
0 & -r & 1 & -\frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\} \nonumber\\
    &-2\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & r & -\frac{r}{\sqrt{2}} \\
0 & 1 & 0 & -\frac{r}{\sqrt{2}} \\
r & 0 & 1 & -\frac{1}{\sqrt{2}} \\
-\frac{r}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\}
\end{align}
Using the facts that
\begin{align}
&\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & r & -\frac{r}{\sqrt{2}} \\
0 & 1 & 0 & -\frac{r}{\sqrt{2}} \\
r & 0 & 1 & -\frac{1}{\sqrt{2}} \\
-\frac{r}{\sqrt{2}} & -\frac{r}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\} \nonumber\\ &+\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & r & \frac{r}{\sqrt{2}} \\
0 & 1 & 0 & \frac{r}{\sqrt{2}} \\
r & 0 & 1 & \frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\} \nonumber\\
=&\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k};\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & r \\
0 & r & 1
\end{pmatrix}\right\}
\end{align}
and
\begin{align}
&\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k};\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & r \\
0 & r & 1
\end{pmatrix}\right\}+\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k};\begin{pmatrix}
1 & 0 & -r \\
0 & 1 & 0 \\
-r & 0 & 1
\end{pmatrix} \right\} \nonumber\\
=&\Phi_{2}(-\Delta_{j}^{1},\Delta_{j}^{2};0)
=\Phi(-\Delta_{j}^{1})\Phi(\Delta_{j}^{2})
\end{align}
So that,
\begin{align}
    F_{NT}(r;\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k})= & -2\Phi(-\Delta_{j}^{1})\Phi(\Delta_{j}^{2}) \nonumber\\
    &+2\Phi_{3}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},\Delta_{k};\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & r \\
0 & r & 1
\end{pmatrix}\right\}\nonumber \\
    & +2\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & 0 & \frac{r}{\sqrt{2}} \\
0 & 1 & -r & \frac{r}{\sqrt{2}} \\
0 & -r & 1 & -\frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & -\frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\} \nonumber\\
    &+2\Phi_{4}\left\{-\Delta_{j}^{1},\Delta_{j}^{2},-\Delta_{k},0;\begin{pmatrix}
1 & 0 & r & \frac{r}{\sqrt{2}} \\
0 & 1 & 0 & \frac{r}{\sqrt{2}} \\
r & 0 & 1 & \frac{1}{\sqrt{2}} \\
\frac{r}{\sqrt{2}} & \frac{r}{\sqrt{2}} & \frac{1}{\sqrt{2}} & 1
\end{pmatrix}\right\}
\end{align}

## Derivation for approximate bound for ternary/continuous case

Let $n_{0x}=\sum_{i=1}^{n_x}I(x_{i}=0)$, $n_{2x}=\sum_{i=1}^{n_x}I(x_{i}=2)$, $\pi_{0x}=\frac{n_{0x}}{n_{x}}$ and $\pi_{2x}=\frac{n_{2x}}{n_{x}}$, then
\begin{align}
    |\tau(\mathbf{x})|\leq & \frac{n_{0x}(n-n_{0x})+n_{2x}(n-n_{0x}-n_{2x})}{\begin{pmatrix} n \\ 2 \end{pmatrix}} \nonumber\\
    = & 2\{\frac{n_{0x}}{n-1}-(\frac{n_{0x}}{n})(\frac{n_{0x}}{n-1})+\frac{n_{2x}}{n-1}-(\frac{n_{2x}}{n})(\frac{n_{0x}}{n-1})-(\frac{n_{2x}}{n})(\frac{n_{2x}}{n-1})\} \nonumber\\
    \approx & 2\{\frac{n_{0x}}{n}-(\frac{n_{0x}}{n})^2+\frac{n_{2x}}{n}-(\frac{n_{2x}}{n})(\frac{n_{0x}}{n})-(\frac{n_{2x}}{n})^2\} \nonumber\\
    = & 2\{\pi_{0x}(1-\pi_{0x})+\pi_{2x}(1-\pi_{0x}-\pi_{2x})\}
\end{align}

## Approximate bound for ternary/binary case and ternary/ternary case

Combine NC and BC case, we get NB case. So does NN case. 

## Derivation for approximate bound for ternary/truncated case

Derivation for approximate bound for ternary truncated case: Let $\mathbf{x}\in\mathcal{R}^{n}$ and $\mathbf{y}\in\mathcal{R}^{n}$ be the observed $n$ realizations of ternary and truncated variables, respectively. Let $n_{0x}=\sum_{i=0}^{n}I(x_{i}=0)$, $\pi_{0x}=\frac{n_{0x}}{n}$, $n_{1x}=\sum_{i=0}^{n}I(x_{i}=1)$, $\pi_{1x}=\frac{n_{1x}}{n}$, $n_{2x}=\sum_{i=0}^{n}I(x_{i}=2)$, $\pi_{2x}=\frac{n_{2x}}{n}$,
$n_{0y}=\sum_{i=0}^{n}I(y_{i}=0)$, $\pi_{0y}=\frac{n_{0y}}{n}$, $n_{0x0y}=\sum_{i=0}^{n}I(x_{i}=0 \;\& \; y_{i}=0)$, $n_{1x0y}=\sum_{i=0}^{n}I(x_{i}=1 \;\& \; y_{i}=0)$ and
$n_{2x0y}=\sum_{i=0}^{n}I(x_{i}=2 \;\& \; y_{i}=0)$ then
\begin{align}
    |\tau(\mathbf{x}, \mathbf{y})|\leq &
    \frac{\begin{pmatrix}n \\ 2\end{pmatrix}-\begin{pmatrix}n_{0x} \\ 2\end{pmatrix}-\begin{pmatrix}n_{1x} \\ 2\end{pmatrix}-\begin{pmatrix} n_{2x} \\ 2 \end{pmatrix}-\begin{pmatrix}n_{0y} \\ 2\end{pmatrix}+\begin{pmatrix}n_{0x0y} \\ 2 \end{pmatrix}+\begin{pmatrix}n_{1x0y} \\ 2\end{pmatrix}+\begin{pmatrix}n_{2x0y} \\ 2\end{pmatrix}}{\begin{pmatrix}n \\ 2\end{pmatrix}} \nonumber
\end{align}
Since $n_{0x0y}\leq\min(n_{0x},n_{0y})$, $n_{1x0y}\leq\min(n_{1x},n_{0y})$ and $n_{2x0y}\leq\min(n_{2x},n_{0y})$ we obtain
\begin{align}
     |\tau(\mathbf{x}, \mathbf{y})|\leq &
    \frac{\begin{pmatrix}n \\ 2\end{pmatrix}-\begin{pmatrix}n_{0x} \\ 2\end{pmatrix}-\begin{pmatrix}n_{1x} \\ 2\end{pmatrix}-\begin{pmatrix} n_{2x} \\ 2 \end{pmatrix}-\begin{pmatrix}n_{0y} \\ 2\end{pmatrix}}{\begin{pmatrix}n \\ 2\end{pmatrix}} \nonumber\\
    & +  \frac{\begin{pmatrix}\min(n_{0x},n_{0y}) \\ 2 \end{pmatrix}+\begin{pmatrix}\min(n_{1x},n_{0y}) \\ 2\end{pmatrix}+\begin{pmatrix}\min(n_{2x},n_{0y}) \\ 2\end{pmatrix}}{\begin{pmatrix}n \\ 2\end{pmatrix}} \nonumber\\
    \leq & \frac{\begin{pmatrix}n \\ 2\end{pmatrix}-\begin{pmatrix}\max(n_{0x},n_{1x},n_{2x},n_{0y}) \\ 2\end{pmatrix}}{\begin{pmatrix}n \\ 2\end{pmatrix}} \nonumber\\
    \leq & 1-\frac{\max(n_{0x},n_{1x},n_{2x},n_{0y})(\max(n_{0x},n_{1x},n_{2x},n_{0y})-1)}{n(n-1)} \nonumber\\
    \approx & 1-(\frac{\max(n_{0x},n_{1x},n_{2x},n_{0y})}{n})^{2} \nonumber\\
    =& 1-\{\max(\pi_{0x},\pi_{1x},\pi_{2x},\pi_{0y})\}^{2} \nonumber\\
    =& 1-\{\max(\pi_{0x},(1-\pi_{0x}-\pi_{2x}),\pi_{2x},\pi_{0y})\}^{2}
\end{align}

# References


