---
title: "latentcor"
author: "Mingze Huang, Irina Gaynanova, Christian L. Müller"
date: "`r Sys.Date()`"
bibliography: latentcor.bib
output: rmarkdown::html_vignette
nocite: |
  @filzmoser2021pcapp
vignette: >
  %\VignetteIndexEntry{latentcor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
library(knitcitations)

bibliography()
write.bibtex(citation("pcaPP"), file = "latentcor.bib")
```
Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.[@filzmoser2021pcapp]

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. `r citep(citation("pcaPP"))`

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```


## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```




bla 

# References


